To be FAIR: Theory Specification Needs an Update

### 2025-06-07 action_letter.Rmd
update 2025-06-07 12:47:56.764452

As methodologists, our focus in this paper was on "procedures" rather than policies or documentation. Our goal was to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.

We agree with the Editor that there is no guarantee that *"these things are crucial to scientific advance in psychology"*. We have made an effort to clearly outline the "affordances" of FAIR theory (what does it make possible), but whether or not these affordances will result in actual scientific advancements remains to be seen.

Compare FAIR theory to preregistration: it, too, was introduced first as a good idea, and now - a decade later - empirical studies are being conducted to investigate its effects.

@@

@@

@@

@@

gonzalezExtrinsicConvergentValidity2021

@@

We tried to make this suggested change, although there is still some heterogeneity because our different institutions have different preferred / official ways to write the affiliation.

@@

>>>>

As methodologists, our focus in this paper is on "procedures", rather than policies or documentation. Our goal is to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.

We agree with the Editor that there is no guarantee that *"these things are crucial to scientific advance in psychology"*. We have, however, provided reasoned arguments to support the "affordances" of FAIR theory: it *makes possible* particular scientific advancements, even if it does not guarantee them.

Compare FAIR theory to preregistration: it, too, was introduced first as a good idea; now - a decade later - empirical studies are being conducted to investigate its effects.

@@

We removed the "cf.", although we couldn't find mention of it in the APA7 style guide. We updated the citation scheme to APA7, so three or more authors will be shortened to "et al.".

@@

Thank you for pointing this out; we have fixed the formatting.

@@

We have added the page numbers to all direct quotations.

@@

We have done as requested; this section now reads:

`r get_revision("jinglejangle")`

@@

We have tried to make this requested change, although there is some residual heterogeneity because our institutions have different preferred / mandated ways to write the affiliation.

@@

We've updated the reference style to APA7, but note that some journals have fixed the case of specific words in titles (e.g., "FAIR" is fixed to capital letters). These remain unchanged. We defer to the Editor to indicate whether we should change this, or whether it will be addressed in the typesetting process.

### 2025-06-07 manuscript.Rmd
update 2025-06-07 12:47:56.764452

abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
keywords          : "fairtheory, meta science, theory formation, open science"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "fair_theory.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, noextraspace"
urlcolor          : blue
output            :
  papaja::apa6_pdf: default
  md_document: default
header-includes:
  - \usepackage{longtable, setspace, subfig}
  - \raggedbottom
---

@@

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:

Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
<!-- Consider, for instance, a research scientist working in a highly competitive laboratory.
Under the original definition, their intrinsic motivation would be limited to the joy of discovery alone.
Yet, under the expanded framework, factors that might initially appear extrinsic can be reconceptualized as intrinsic: securing grant funding enhances their autonomy by granting them the freedom to pursue innovative ideas; grappling with deadlines bolsters their sense of competence; and praise from their peers fulfills their need for relatedness. -->
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
The expanded definition thus enables SDT to absorb potentially falsifying evidence.

@@

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].
The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory are abundant, and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our suggested approach is not limited to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on their website <www.theorymaps.org> [@grayHowMapTheory2017].
Similarly, [PsychoModels](http://psychomodels.org/) seeks to inventorize theories and models in psychology [@vandongenPsychoModelsMathematicalComputational2025].
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

A second impediment to Accessibility is more indirect
and pertains to a theory's intelligibility to those with practical Access.
It has been proposed that good theories have the property of "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. 
At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
Shared background knowledge can come from paradigms held by members of a scientific community [@kuhnStructureScientificRevolutions2009],
from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
At the same time, it is important to acknowledge the *indeterminacy of translation*^[Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
has multiple alternative translations, with no objective means of choosing the correct one.],
which implies that it is not possible to remove *all* ambiguity when communicating an idea [@quineReasonsIndeterminacyTranslation1970].
This places a theoretical upper bound on theories' ability to be understood.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).

@@

Recent work points out that Interoperability is not an all-or-nothing property.
The concept of X-Interoperability was introduced to answer the question: *Interoperable for what?*
X-Interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-Interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
This revised definition makes it possible to outline a theory's affordances in terms of X-Interoperability.
For example, a FAIR theory may be X-Interoperable for deriving testable hypotheses,
or for the purpose of selecting relevant control variables,
or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
we see how each of these properties incurs certain affordances in terms of X-Interoperability (Table \@ref(tab:tabmeehl)).

@@

If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologyFoundationsInference1969].
In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.
Students ask questions such as "Who owns a theory?",
and "Who determines how a theory may be reused or changed?".
These questions imply a norm against modifying theory without its author's consent, reminiscent of the aforementioned problem of dependency on the author.

Licensing theories for reuse unambiguously answers these questions,
with the caveats that legislation may vary across contexts and jurisdictions, and that this paper does not constitute legal advice.
Two considerations are important when determining what license is appropriate for theory.
A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy [@bently2010copyright].
It explicitly does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
but not to the underlying theoretical idea.
It thus seems that theories expressed in prose or depicted visually - in other words, that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
<!-- If copyright limits Reusability and does not cover ideas in their purest form (like formal theories), -->
<!-- then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection. -->
A second consideration is that academic research is covered under "fair use" exemptions to copyright.
<!-- so copyright would pose few restrictions to Reusability in scholarly communication. -->
Given these two considerations - that copyright does not protect ideas in their purest form and that academic use offers exemptions to copyright - it may be counterproductive and possibly misleading to adopt a license that assumes copyright protection to theories.
For psychological theories without commercial aspects, we suggest using a licence that explicitly waives copyright and encourages Reusability,
such as CC0 (no rights reserved).

@@

Several authors have reinterpreted De Groot's empirical cycle.
An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
For example, the DOT graph below implements Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle.
First, notice that the phases of the cycle have been renamed.
This change was not described in the paper.
If we assume that the labels are meant to illustrate the phases,
not substantially change the ontology,
then we can represent it by adding labels to the original DOT graph.
These labels suggest a focus on empirical psychology that was not present in De Groot's version.
Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.

>>>>

abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
keywords          : "fairtheory, meta science, theory formation, open science"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "fair_theory.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, noextraspace"
urlcolor          : blue
csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            :
  papaja::apa6_pdf: default
  md_document: default
header-includes:
  - \usepackage{longtable, setspace, subfig}
  - \raggedbottom
---

@@

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].

Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
rendering them immune to falsification.
As an example, consider "self-determination theory" (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness [@ryanSelfdeterminationTheoryFacilitation2000].
<!-- Consider, for instance, a research scientist working in a highly competitive laboratory.
Under the original definition, their intrinsic motivation would be limited to the joy of discovery alone.
Yet, under the expanded framework, factors that might initially appear extrinsic can be reconceptualized as intrinsic: securing grant funding enhances their autonomy by granting them the freedom to pursue innovative ideas; grappling with deadlines bolsters their sense of competence; and praise from their peers fulfills their need for relatedness. -->
The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
Under the original definition, one would hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
The expanded definition thus enables SDT to absorb potentially falsifying evidence.

@@

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978, p. 1].
The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory are abundant, and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our suggested approach is not limited to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p. 336].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on [their website](https://www.theorymaps.org) [@grayHowMapTheory2017].
Similarly, [PsychoModels](http://psychomodels.org/) seeks to inventorize theories and models in psychology [@vandongenPsychoModelsMathematicalComputational2025].
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

A second impediment to Accessibility is more indirect
and pertains to a theory's intelligibility to those with practical Access.
It has been proposed that good theories have the property of "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024, p. 1]. 
At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
Shared background knowledge can come from paradigms held by members of a scientific community [@kuhnStructureScientificRevolutions2009],
from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
At the same time, it is important to acknowledge the *indeterminacy of translation*^[Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
has multiple alternative translations, with no objective means of choosing the correct one.],
which implies that it is not possible to remove *all* ambiguity when communicating an idea [@quineReasonsIndeterminacyTranslation1970].
This places a theoretical upper bound on theories' ability to be understood.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016, p. 2].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).

@@

Recent work points out that Interoperability is not an all-or-nothing property.
The concept of X-Interoperability was introduced to answer the question: *Interoperable for what?*
X-Interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-Interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024, p. 5].
This revised definition makes it possible to outline a theory's affordances in terms of X-Interoperability.
For example, a FAIR theory may be X-Interoperable for deriving testable hypotheses,
or for the purpose of selecting relevant control variables,
or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
we see how each of these properties incurs certain affordances in terms of X-Interoperability (Table \@ref(tab:tabmeehl)).

@@

If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008, p. 1].
As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologyFoundationsInference1969].
In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.
Students ask questions such as "Who owns a theory?",
and "Who determines how a theory may be reused or changed?".
These questions imply a norm against modifying theory without its author's consent, reminiscent of the aforementioned problem of dependency on the author.

Licensing theories for reuse unambiguously answers these questions,
with the caveats that legislation may vary across contexts and jurisdictions, and that this paper does not constitute legal advice.
Two considerations are important when determining what license is appropriate for theory.
A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy [@bently2010copyright].
Copyright does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"* (Section 102, Copyright Act).
Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
but not to the underlying theoretical idea.
It thus seems that theories expressed in prose or depicted visually - in other words, that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
<!-- If copyright limits Reusability and does not cover ideas in their purest form (like formal theories), -->
<!-- then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection. -->
A second consideration is that academic research is covered under "fair use" exemptions to copyright.
<!-- so copyright would pose few restrictions to Reusability in scholarly communication. -->
Given these two considerations - that copyright does not protect ideas in their purest form and that academic use offers exemptions to copyright - it may be counterproductive and possibly misleading to adopt a license that assumes copyright protection to theories.
For psychological theories without commercial aspects, we suggest using a licence that explicitly waives copyright and encourages Reusability,
such as CC0 (no rights reserved).

@@

Several authors have reinterpreted De Groot's empirical cycle.
An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
For example, the DOT graph below implements Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle.
First, notice that the phases of the cycle have been renamed.
This change was not described in the paper.
If we assume that the labels are meant to illustrate the phases,
not substantially change the ontology,
then we can represent it by adding labels to the original DOT graph.
These labels suggest a focus on empirical psychology that was not present in De Groot's version.
Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"* (p. 423).
The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
